\includegraphics*[width=0.4\textwidth]{hpi_logo_2017.eps}\\ \vspace{24pt} Genetic Doodle Jump

### 2018-03-31 paper/paper.tex
small improvements

We did something similar in our research seminar, but with a less complex game: Doodle Jump. The player of this video game has to jump on randomly located platforms in order to get higher and thus, get a higher score. In our approach we use a genetic algorithm in combination with neural networks for the machine learning part. An artificial neural net (ANN) is a machine learning algorithm inspired by nature. It consists of neurons, which calculate an output based on an input similar to a biological nerve cell. Those neurons are connected with each other to enable an information flow within the network. In order to improve the performance of an ANN, many parameters --- for example the biases of the neurons or the weight of connections --- can be tuned.

At this point we make use of the genetic algorithm. This algorithm is modelled on biological evolution. There are many individuals which form a population. Applied to our implementation, those are the different neural nets playing the game. After all players died, we select the best of them and perform crossover and mutation like in nature on those. Hence, we create a new generation, which is supposed to be better than the previous one. In this way, we always create better and better neural networks.

@@

Shohet uses a Q-Learning algorithm, which is model-free. This technique depends on states. Depending on the current state an action will be calculated in order to go into the next state. Based on the impact of the action a reward will be awarded. If this action has a negative effect, for example the player falls down and dies, the evaluation is bad. Otherwise, the it will be positive. Based on those rewards, the algorithm learns when to use which actions in order to become better.

@@

The video game Doodle Jump is developed for mobile devices by Lima Sky\footnote{\url{http://www.limasky.com/}}. The player --- the so called \quot{doodler} --- has to jump on platforms in order to get a higher score. There are four different types of platforms:
\begin{enumerate}
    \item Regular ones, which neither move nor break
    \item Horizontally moving platforms
    \item Vertically moving platforms
    \item Vanishable ones, which disappear after jumping on it once
    \item Breaking platforms, where the doodler can not jump on
\end{enumerate}
The platforms are pseudo randomly located on the map. Pseudo randomly located means, that there will always be at least one platform, which is reachable from the previous one and non-breakable (See Fig. \ref{abb:doodlejumpgame}). In other words, there is always a way to reach an infinite score. However, the higher the achieved score, the more difficult the game becomes. The difficulty is reached via higher distances between the platforms and the increased usage of moving and vanishable platforms.

@@

Since we do not want to perform the deep learning part on a mobile device or within an emulator, we decided to use an implementation, which runs in a browser. We are using \quot{HTML5 Doodle Jump} by Cong Nguyen\footnote{\url{https://github.com/ntcnet83/html5-doodle-jump}}, which we found on github\footnote{\url{https://github.com/}}. This custom version of the is written in HTML5 and JavaScript. The doodler can be control via left and right arrow keys. In contrast to the original implementation, this version of the game contains less features, which makes the game simpler at the end:
\begin{itemize}
\item Except springs, there are no other bonuses.
\item There are no enemies. Hence, the only way to die is falling down.
\item The feature shooting is not necessary and therefore not available.
\item The game does not have vertically moving platforms.
\item The difficulty changes after a certain score, but only through the usage of more moving and vanishable platforms. The distances between the platforms stay the same.
\end{itemize}

Our machine learning part runs in the browser as well. We use the \textit{Synaptic} library\footnote{\url{https://github.com/cazala/synaptic}} in order to create our neural networks. This library provides a high level interface to simply initialize a perceptron. Furthermore, adapted the genetic algorithm implemented in the \textit{Genetic.js} library\footnote{\url{https://github.com/subprotocol/genetic-js}}. Both libraries are written in JavaScript.

@@

Multiple neural networks are playing concurrently and each neural network plays two games with a different platform arrangement. Those numbers (i.e. number of neural networks and number of games played by one neural network) can be configured in our application. However, the screenshot only shows two of those 16 neural networks. Nevertheless, the user can scroll down in the UI to inspect each game which is running at this point. The UI also has buttons to start, stop and resume the games.

@@

We wanted to add some outputs to inform about the overall learning progress. The outputs ``best score'', ``best fitness'' and ``best score per generation'' are displayed as values. These values help to understand if the neural networks are still improving. If the scores are not increasing, this is a good point to debug what the problem is. For example, the doodler could target platforms that are harder to reach than other platforms. Then the user could disable the speed up and he could precisely inspect situations where this happens. During our project this kind of debugging helped a few times. Mostly, we had to adjust the fitness function or input parameters for the genetic algorithm to solve such problems. Some of these details are explained in section \ref{sec:id}.

@@

The input of the neural network is the information which a player has of the game at the current state. Section \ref{sec:ip} describes which input parameters were used. The output of the neural network is the action which the player is going to make. The actions of the Doodle Jump implementation are very simple. The player can either move left or move right or stop moving. The calculated values in the three output neurons are in the range from 0 to 1. The maximum value of these three outputs is used to determine the actual action for the doodler. When the outputs of the network for a given state would be $moveLeft: 0.21, moveRight: 0.07, stopMoving: 0.19$ the doodler would move to the left.

@@

Figure \ref{abb:nl} shows that we have 12 input parameters. The first two parameters are the ``x-velocity'' and ``y-velocity'' of the doodler. ``x-velocity'' tells how fast the doodler is moving horizontally. The longer the doodler keeps moving in one direction the faster it is moving in this direction (until it reaches speed limits). A positive ``x-velocity'' value means that the doodler is moving to the right and a negative value means that it is moving to the left. There is no horizontal movement if ``x-velocity'' is 0.

The ``y-velocity'' works similar with the vertical speed. The ``y-velocity'' has the highest value right after jumping on a platform. Then it starts decreasing until the doodler starts moving down and the ``y-velocity'' becomes negative.

@@

To clarify how the input parameters are calculated in a specific game state we created the example in figure \ref{abb:ip}. The ``x-velocity'' is $0.2$. So the doodler has almost no horizontal speed. The ``y-velocity'' is $7.8$ which is very high because the doodler just jumped on the blue platform. The next two platforms (the white and the green platform) are used for the input parameters. The white platform is vanishable and the green platform is a regular one (see section \ref{sec:og}). The center of the doodler and of the platforms are used to calculate the relative distances. Regarding the ``relative x-distance'' to the white platform it is shorter to pass through the right wall. So we always check which direction is shorter. In the end the genetic algorithm does not have to know where the walls are. The relative distances to the white platform are $192px$ and $12px$. The relative distances to the green platform are $-80px$ and $67px$. The vertical distances can also be negative when the doodler gets above the platforms. This are the 12 input parameters as an array: $[0.2, 7.8, 193, 12, 0, 0, 0, -80, 67, 0, 0, 0]$.

``has spring'', ``is breakable'' and ``is moving'' are either 0 or 1. Because both input platforms have no spring, are not breakable and are not moving these values are 0 in the example. Since the doodler should jump on each platform only once (even the regular platforms) we do not distinguish vanishable (white) platforms and regular (green) platforms.

@@

Doodle Jump is a game where each level is generated randomly. We want to end up with a network that is able to play any level well, regardless of the specific generated layout. The performance of any one network on any specific level may not necessarily be indicative of the networks ability to play the game overall. Therefore, each network plays several levels simultaneously, and each level is taken into account when evaluating the performance.

@@

Beyond that, there are many other fitness function one could think of. For example, at the very beginning of the optimization process, many candidates might do nothing at all, while others move a little bit into the direction of the next platform, but not far enough to actually land on it. All candidates would thus achieve a score of zero, and the primitive fitness function would be unable to discriminate between them. Another fitness function, which rewards candidates that got closer to the platform they failed to jump on might speed up the learning process in the first few generations. It would also be helpful later on, when different candidates might fail at the same platform, thus achieving an equal score. However, such a fitness function would include prior knowledge about how to play the game well. We want to avoid using such fitness functions, since for other problems, prior knowledge might not be available, or might not lead to the optimal solution if available.

@@

All functions lead to a learning where to expected score is rising close to linearly. Score divided by time leads to the slowest leaning, since it essentially just measures the (nearly constant) speed in which any candidate is traveling upwards, so the fitness is nearly constant as well. All other functions involving time perform better than just taking the score. $fitness_3$ is one of the simplest ways of rewarding for gained score while penalizing for taken time, whereas $fitness_5$ puts emphasis on score and only uses time to break ties between two candidates that obtained the same score.

@@

After the fitness of each candidate has been determined, the $n$ top candidates are selected to generate offspring that form the next generation. Here, $n$ is a parameter that can either be chosen or dynamically adapted during the algorithm. Choosing a higher $n$ leads to more diverse offspring and the possibility to find small local optima that might not be found otherwise, but as their parents are worse, usually they can be expected to be worse as well.

@@

We perform mutation on the biases of neurons in the same we mutate the weights of the connections. First, we iterate over all elements. Each time a random float between zero and one is created If this random value is lower then the mutation rate we mutate the element, otherwise not. Then, the bias or weight is changed to the value of the formula $1 + ((Math.random() - 0.5) * 3 + (Math.random() - 0.5))$, which is at least $-1$ and $3$ at maximum.
\subsection{Retaining best candidates}

After all games died with one generation we select the top players of this run and reuse them without any modification in the next generation. The amount of retained candidates depends on the performance of the current run compared with the previous one, similar to the mutation rate. However, the number of players kept is limited to the range of three to five out of sixteen. The reason, why we have such a complex algorithm is, that we want to a high amount of good players and throw away a large number of bad ones. In order to explain it in detail we use an example which is depicted in graphic \ref{abb:retaining}. It shows the players of three generations sorted by fitness.

@@

The second generation performs a little worse than the previous one. Hence, we will keep less candidates. First, we compare the top five players with the previous run. In this concrete example only two players are better than the last best five. We then also select the third player, because this is the minimum number of retained candidates. The main idea behind this lower limit is, that we do not want to throw away to many information.

@@

Initially when we started working on the project, there existed a huge local optimum where candidates that would choose to accelerate to the right (or left) constantly would quickly go so fast that they were able to hit a good number of platforms before falling down, therefore easily beating any initial candidates that tried to work with the input data in any way. This behavior did not resemble how a human would approach the game, nor was it visually appealing. It also was not a sophisticated strategy (as any information about the surroundings of the doodler would just be discarded) and had a rather low ceiling in terms of how many points could be achieved that way.

In order to counteract this, we decided to modify the game in a way that any player touching the walls would die, instead of reappearing on the other side of the level. This lead to an interesting effect in combination with the third fitness function, $score - time$. Score is awarded only if at least one platform is being reached. Time, on the other hand, starts counting immediately. Therefore, a candidate that would slowly move into the direction of the first platform, but not fast enough to reach it before he would die, would get a negative fitness. Another candidate that always goes right, therefore hitting the right wall as soon as possible, also gets a negative fitness, but due to the speed of his suicide, he still gets a better fitness than the first candidate. Ultimately, after a few generations, all candidates had learn to commit suicide as quickly as possible. We could resolve this problem by treating every negative fitness as equal.

>>>>

We did something similar in our research seminar, but with a less complex game: Doodle Jump. The player of this video game has to jump on randomly located platforms in order to get higher and thus, get a higher score. In our approach we use a genetic algorithm in combination with neural networks for the machine learning part. An artificial neural net (ANN) is a machine learning algorithm inspired by nature. It consists of neurons, which calculate an output based on an input, similar to a biological nerve cell. Those neurons are connected with each other to enable an information flow within the network. In order to improve the performance of an ANN, many parameters --- for example the biases of the neurons or the weight of connections --- can be tuned.

At this point we make use of the genetic algorithm. This algorithm is modelled on biological evolution. There are many individuals which form a population. Applied to our implementation, those are the different neural nets playing the game. After all players died, we select the best of them and perform crossover and mutation on those. Hence, we create a new generation, which is supposed to be better than the previous one. In this way, we always create better and better neural networks.

@@

Shohet uses a Q-Learning algorithm, which is model-free. This technique depends on states. Depending on the current state an action will be calculated in order to go into the next state. Based on the impact of the action a reward will be awarded. If this action has a negative effect, for example the player falls down and dies, the evaluation is bad. Otherwise, it will be positive. Based on those rewards, the algorithm learns when to use which actions in order to become better.

@@

The video game Doodle Jump is developed for mobile devices by Lima Sky\footnote{\url{http://www.limasky.com/}}. The player --- the so called \quot{doodler} --- has to jump on platforms in order to get a higher score. There are five different types of platforms:
\begin{enumerate}
    \item Regular ones, which neither move nor break
    \item Horizontally moving platforms
    \item Vertically moving platforms
    \item Vanishable ones, which disappear after jumping on it once
    \item Breaking platforms, where the doodler can not jump on
\end{enumerate}
The platforms are pseudo-randomly located on the map. Pseudo randomly located means, that there will always be at least one platform, which is reachable from the previous one and non-breakable (See Fig. \ref{abb:doodlejumpgame}). In other words, there is always a way to reach an infinite score. However, the higher the achieved score, the more difficult the game becomes. The difficulty is reached via higher distances between the platforms and the increased usage of moving and vanishable platforms.

@@

Since we do not want to perform the deep learning part on a mobile device or within an emulator, we decided to use an implementation, which runs in a browser. We are using \quot{HTML5 Doodle Jump} by Cong Nguyen\footnote{\url{https://github.com/ntcnet83/html5-doodle-jump}}, which we found on GitHub\footnote{\url{https://github.com/}}. This custom version of the game is written in HTML5 and JavaScript. The doodler can be controlled via left and right arrow keys. In contrast to the original implementation, this version of the game contains less features, which makes the game simpler at the end:
\begin{itemize}
\item Except springs, there are no other bonuses.
\item There are no enemies. Hence, the only way to die is falling down.
\item The feature shooting is not necessary and therefore not available.
\item The game does not have vertically moving platforms.
\item The difficulty changes after a certain score, but only through the usage of more moving and vanishable platforms. The distances between the platforms stays the same.
\end{itemize}

Our machine learning part runs in the browser as well. We use the \textit{Synaptic} library\footnote{\url{https://github.com/cazala/synaptic}} in order to create our neural networks. This library provides a high level interface to simply initialize a perceptron. Furthermore, we adapted the genetic algorithm implemented in the \textit{Genetic.js} library\footnote{\url{https://github.com/subprotocol/genetic-js}}. Both libraries are written in JavaScript.

@@

Multiple neural networks are playing concurrently and each of those plays two games with a different platform arrangement. Those numbers (i.e. number of neural networks and number of games played by one neural network) can be configured in our application. However, the screenshot only shows two of those 16 neural networks. Nevertheless, the user can scroll down in the UI to inspect each game which is running at this point. The UI also has buttons to start, stop and resume the games.

@@

We wanted to add some outputs to inform about the overall learning progress. The outputs ``best score'', ``best fitness'' and ``best score per generation'' are displayed as values. These values help to understand if the neural networks are still improving. If the scores are not increasing, this is a good point to debug what the problem is. For example, the doodler could target platforms that are harder to reach than other platforms. Then the user could disable the speed up and  precisely inspect situations where this happens. During our project this kind of debugging helped a few times. Mostly, we had to adjust the fitness function or input parameters for the genetic algorithm to solve such problems. Some of these details are explained in section \ref{sec:id}.

@@

The input of the neural network is the information which a player has of the game at the current state. Section \ref{sec:ip} describes which input parameters were used. The output of the neural network is the action which the player is going to make. The actions of the Doodle Jump implementation are very simple. The player can either move left or move right or stop moving. The calculated values of the three output neurons are in the range of 0 to 1. The maximum value of these three outputs is used to determine the actual action for the doodler. When the outputs of the network for a given state would be $moveLeft: 0.21, moveRight: 0.07, stopMoving: 0.19$ the doodler would move to the left.

@@

Figure \ref{abb:nl} shows that we have 12 input parameters. The first two parameters are the ``x-velocity'' and ``y-velocity'' of the doodler. ``x-velocity'' tells how fast the doodler is moving horizontally. The longer the doodler keeps moving in one direction the faster it is moving in this direction (until it reaches a certain speed limit). A positive ``x-velocity'' value means that the doodler is moving to the right and a negative value means that it is moving to the left. There is no horizontal movement if ``x-velocity'' is 0.

The ``y-velocity'' works similar to the vertical speed. The ``y-velocity'' has the highest value right after jumping on a platform. Then it starts decreasing until the doodler starts moving down and the ``y-velocity'' becomes negative.

@@

To clarify how the input parameters are calculated in a specific game state we created an example in figure \ref{abb:ip}. The ``x-velocity'' is $0.2$. So the doodler has almost no horizontal speed. The ``y-velocity'' is $7.8$ which is very high because the doodler just jumped on the blue platform. The next two platforms (the white and the green platform) are used for the input parameters. The white platform is vanishable and the green platform is a regular one (see section \ref{sec:og}). The center of the doodler and of the platforms are used to calculate the relative distances. Regarding the ``relative x-distance'' to the white platform it is shorter to pass through the right wall. So we always check which direction is shorter. In the end the genetic algorithm does not have to know where the walls are. The relative distances to the white platform are $192px$ and $12px$. The relative distances to the green platform are $-80px$ and $67px$. The vertical distances can also be negative when the doodler gets above the platforms. These are the 12 input parameters as an array: $[0.2, 7.8, 193, 12, 0, 0, 0, -80, 67, 0, 0, 0]$.

The parameters ``has spring'', ``is breakable'' and ``is moving'' are either 0 or 1. Because both input platforms have no spring, are not breakable and are not moving these values are 0 in the example. Since the doodler should jump on each platform only once (even the regular platforms) we do not distinguish vanishable (white) platforms and regular (green) platforms.

@@

Doodle Jump is a game where each level is generated randomly. We want to end up with a network that is able to play any level well, regardless of the specific generated layout. The performance of any network on any specific level may not necessarily be indicative of the networks ability to play the game overall. Therefore, each network plays several levels simultaneously, and each level is taken into account when evaluating the performance.

@@

Beyond that, there are many other fitness functions one could think of. For example, at the very beginning of the optimization process, many candidates might do nothing at all, while others move a little bit into the direction of the next platform, but not far enough to actually land on it. All candidates would thus achieve a score of zero, and the primitive fitness function would be unable to discriminate between them. Another fitness function, which rewards candidates that got closer to the platform they failed to jump on might speed up the learning process in the first few generations. It would also be helpful later on, when different candidates might fail at the same platform, thus achieving an equal score. However, such a fitness function would include prior knowledge about how to play the game well. We want to avoid using such fitness functions, since for other problems, prior knowledge might not be available, or might not lead to the optimal solution if available.

@@

All functions lead to a learning where the expected score is rising close to linearly. Score divided by time leads to the slowest leaning, since it essentially just measures the (nearly constant) speed in which any candidate is traveling upwards, so the fitness is nearly constant as well. All other functions involving time perform better than just taking the score. $fitness_3$ is one of the simplest ways of rewarding for gained score while penalizing for taken time, whereas $fitness_5$ puts emphasis on score and only uses time to break ties between two candidates that obtained the same score.

@@

After the fitness of each candidate has been determined, the $n$ top candidates are selected to generate offsprings that form the next generation. Here, $n$ is a parameter that can either be chosen or dynamically adapted during the algorithm. Choosing a higher $n$ leads to more diverse offspring and the possibility to find small local optima that might not be found otherwise, but as their parents are worse, usually they can be expected to be worse as well.

@@

We perform mutation on the biases of neurons in the same we mutate the weights of the connections. First, we iterate over all elements. Each time a random float between 0 and 1 is created. If this random value is lower then the mutation rate we mutate the element, otherwise not. Then, the bias or weight is changed to the value of the formula $1 + ((Math.random() - 0.5) * 3 + (Math.random() - 0.5))$, which is at least $-1$ and $3$ at maximum.
\subsection{Retaining best candidates}

After all games died within one generation we select the top players of this run and reuse them without any modification in the next generation. The amount of retained candidates depends on the performance of the current run compared with the previous one --- similar to the mutation rate. However, the number of players kept is limited to the range of three to five out of 16. The reason, why we have such a complex algorithm is, that we want to a high amount of good players and throw away a large number of bad ones. In order to explain it in detail we use an example which is depicted in graphic \ref{abb:retaining}. It shows the best players of three generations sorted by fitness.

@@

The second generation performs a little worse than the previous one. Hence, we will keep less candidates. First, we compare the top five players with the previous run. In this concrete example only two players are better than the last best five. We then also select the third player, because this is the minimum number of retained candidates. The main idea behind this lower limit is, that we do not want to throw away too much information.

@@

Initially when we started working on the project, there existed a huge local optimum where candidates that would choose to accelerate to the right (or left) constantly. The doodler accelerated rapidly and was so fast that it was able to hit a good number of platforms before falling down, therefore easily beating any initial candidates that tried to work with the input data in any way. This behavior did not resemble how a human would approach the game, nor was it visually appealing. Additionally, it was not a sophisticated strategy (as any information about the surroundings of the doodler would just be discarded) and had a rather low ceiling in terms of how many points could be achieved that way.

In order to counteract this, we decided to modify the game in a way that any player touching the walls would die, instead of reappearing on the other side of the level. This lead to an interesting effect in combination with the third fitness function, $score - time$. Score is awarded only if at least one platform is being reached. Time, on the other hand, starts counting immediately. Therefore, a candidate that would slowly move into the direction of the first platform, but not fast enough to reach it before he would die, would get a negative fitness. Another candidate that always goes right, therefore hitting the right wall as soon as possible, also gets a negative fitness, but due to the speed of his suicide, he still gets a better fitness than the first candidate. Ultimately, after a few generations, all candidates learned to commit suicide as quickly as possible. We could resolve this problem by treating every negative fitness as equal.
